Universals of word order result from optimization of grammars for efficient communication

### 2019-02-05 second-draft/paper-names.tex
continue revising

Natural languages vary a lot in the order in which they express information.
Consider Figure~\ref{fig:arabic-japanese-simple}, showing a sentence in Arabic (top) and Japanese (bottom), both translating to `I wrote a letter to a friend.'
Both sentences contain a verb meaning `wrote', a noun expressing  `letter', and a phrase translating to `to a friend'.
In linguistic terminology, `letter' is the object of the verb, whereas the phrases translating as `to a friend' are known as adpositional phrases.
While the two sentences containing words with the same meanings, the order of these words are entirely different in the two languages:
%However, the languages differ in the order in which these words and phrases are arranged: 
In Arabic, the verb stands at the beginning, followed by both the object and the adpositional phrase.
In Japanese, the verb stands at the end; object and adpositional phrase precede it.
What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb.
It turns out that this is a deep and stable regularity:
Languages ordering the objects before (Japanese) or after (Arabic) the verb are approximately equally common around the world.
However, the vast majority of natural languages place adpositional phrases on the same side of the verb as the object:
In a sample of about 1,100 languages from five contents (CITE Dryer WALS 95a), less than 7 \% of languages provide clear exceptions to this generalization. %, over 90 \% of languages satisfy this generalization .

This generalization falls in a group of language universals that were originally documented by Greenberg \cite{greenberg1963universals}, known as \key{word order correlations}:
These describe correlations between the relative positions of different types of expressions across languages.
The example above documents that the position of the object (`letter') relative to the verb is \key{correlated} with the position of adpositional phrases (`to a friend') relative to the verb.
Greenberg originally worked with 30 languages; the correlation universals have since been confirmed on the basis of much larger samples of languages.
The authoritative study by Dryer \cite{dryer1992greenbergian} draws on 625 languages and documents fifteen such correlations.
%For instance, the relative order of objects and nouns also strongly determines the positions of genitives, auxiliaries, and other

These universals are supported by languages on all five continents, and they are valid across

These cross-linguistic patterns call for an explanation, and the problem of explaining the word order correlations has inspired a long tradition of research.

A prominent line of research has argued that universals arise for \key{functional} reasons: that is, because they make human communication and language processing maximally efficient, and regularities across languages hold because these efficiency constraints are rooted in general principles of communication and cognition (e.g., \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language}).
Under this view, the various human languages represent multiple solutions to the problem of efficient information transfer given human cognitive constraints.

@@

\begin{table*}
	\begin{center}
\begin{tabular}{|ll|l|l|l|ll|l|llllllll}
	\hline
	%&
	\multicolumn{2}{|c|}{Correlates with...}   &         \multirow{2}{*}{Real}   & \multirow{2}{*}{DepL}  & \multirow{2}{*}{Pred}   &  \multirow{2}{*}{Pars} &  \multirow{2}{*}{Efficiency}  \\ 
	verb & object     & & & &  & \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
adposition    &    NP       & \\
	to            & a friend \\
copula    &    NP        &     \\
	is        & a friend \\
auxiliary    &    VP       &   \\
	has          & written \\
noun    &    genitive      &  \\
TODO &  \\
noun    &    relative clause       &  \\
complementizer    &    S        & \\
verb    &    PP     &    \\
want    &    VP      &    \\
verb    &    subject    &  \\
verb    &    manner adverb     & \\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
    \multicolumn{6}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}
\end{tabular}
	\end{center}
	\caption{Greenbergian Correlations. Following \cite{dryer1992greenbergian}, each correlation is stated in terms of a pair of a `verb patterner' and an `object patterner', whose relative order correlate with that of verbs and objects.
}\label{table:results-dryer}
\end{table*}

>>>>

Natural languages vary a lot in the order in which they express information.
Consider Figure~\ref{fig:arabic-japanese-simple}, showing a sentence in Arabic (top) and Japanese (bottom), both translating to `I wrote a letter to a friend.'
Both sentences contain a verb meaning `wrote', a noun expressing  `letter', and a phrase translating to `to a friend'.
In linguistic terminology, `letter' is the object of the verb, whereas the phrases translating as `to a friend' are known as adpositional phrases.
While the two sentences containing words with the same meanings, the order of these words are entirely different in the two languages:
%However, the languages differ in the order in which these words and phrases are arranged: 
In Arabic, the verb stands at the beginning, followed by both the object and the adpositional phrase.
`To' is expressed by a \emph{preposition}, so named because it precedes the word expressing `friend'. %The adpositional phrase starts with the adposition.
In Japanese, the verb stands at the end; object and adpositional phrase precede it; `to' is expressed by a \textit{postposition}, so named because it follows the word denoting `friend'.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb, and that the objects of adpositions and verbs stand on the sam side of their heads.
%What both languages have in common is that the object and the adpositional phrase stand on the same side of the verb.
%It turns out that these are deep and stable regularities:
It turns out that this variation reflects a deep and stable regularity:
While languages ordering the objects before (Japanese) or after (Arabic) the verb are approximately equally common around the world,
this is strongly correlated with the occurrence of pre- or postpositions:
Languages ordering their objects the way Japanese does, have postpositions; languages ordering them as Arabic does have prepositions.
%However, the vast majority of natural languages place adpositional phrases on the same side of the verb as the object:
This correlation has an extremely strong empirical basis:
In a sample of about 1,100 languages from five contents (CITE Dryer WALS 95a), less than 7 \% of languages provide clear exceptions to this generalization. %, over 90 \% of languages satisfy this generalization .

@@

This generalization falls in a group of language universals that were originally documented by Greenberg \cite{greenberg1963universals}, known as \key{word order correlations}:
These describe correlations between the relative positions of different types of expressions across languages.
The example above documents that the position of the object (`letter') relative to the verb is \key{correlated} with the position of the adposition (`to'). %adpositional phrases (`to a friend') relative to the verb.
Greenberg originally worked with 30 languages; the correlation universals have since been confirmed on the basis of much larger samples of languages.
The authoritative study by Dryer \cite{dryer1992greenbergian} draws on 625 languages and documents fifteen such correlations.
%For instance, the relative order of objects and nouns also strongly determines the positions of genitives, auxiliaries, and other

%These universals are supported by languages on all five continents, cutting across language families.

%These cross-linguistic patterns call for an explanation, and 
Explaining these patterns has been an important aim of linguistic research since Greenberg's seminal study.
%and the problem of explaining the word order correlations has inspired a long tradition of research.
A prominent line of research has argued that universals arise for \key{functional} reasons: that is, because they make human communication and language processing maximally efficient, and regularities across languages hold because these efficiency constraints are rooted in general principles of communication and cognition (e.g., \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language}).
Under this view, the various human languages represent multiple solutions to the problem of efficient information transfer given human cognitive constraints.

We formalize efficiency of language as a tradeoff between minimizing the

We go back to early functional work.
\cite{zipf1949human} argued that language optimizes a tradeoff between speaker effort (Force of Unification) and listener effort (Force of Diversification).
According to Zipf, a language that minimizes speaker effort should reduce the number of different utterances to make production as easy as possible.
Conversely, to minimize listener effort, the language should provide different utterances for different meanings, so that the listeenr can unambiguously identify the meaning from the utterance.
Minimizing speaker and listener represent two opposing forces:
A pressure to reduce the complexity of the language makes the utterances in a language more homogenous.
A pressure to reduce ambiguity makes utterances more heterogeneous, so they can indicate different meanings.
The idea that language results from the tension between these two pressures has a long and fruitful history.
\cite{horn-toward-1984}, RSA, Color naming

This idea has been shown to account for phenomena such as pragmatic inference (CITE) and color naming (CITE).

TODO somewhere start calling this Efficiency

TODO explain why surprisal. some production/speaker work. (plan reuse)

%In restricted settings such as modeling reference games and the distribution of meanings in certain classes of words, these ideas have received computational and information-theoretic formalizations, showing that they successfully account for phenomena such as pragmatic inference (CITE frank and goodman) and color naming (CITE).
%Even earlier than Zipf, \cite{gabelentz1901sprachwissenschaft} ...
%This theory blabla Horn RSA
%Prior computational implementations of this idea have mostly been in the restricted small-scale settings, such as simple reference games.

In this work, we show computationally that efficiency optimization accounts for the Greenbergian word order correlations.
That is, we show that word orders in natural language

%\section{DLM}

%
%\begin{figure}
%    \centering
%    \includegraphics[width=0.45\textwidth]{figures/letter-options.png}
%	\caption{The DLM theory of word order universals: We show how four possible grammars can order the given dependency structure. In this example, Grammar 1 achieves the lowest overall dependency length. Indeed, this is the }
%	\label{fig:pareto-plane}
%\end{figure}
%
%

\begin{table*}
	\begin{center}
\begin{tabular}{|ll|l|l|l|ll|l|llllllll}
	\hline
	%&
	\multicolumn{2}{|c|}{Correlates with...}   &         \multirow{2}{*}{Real}   & \multirow{2}{*}{DepL}  & \multirow{2}{*}{Pred}   &  \multirow{2}{*}{Pars} &  \multirow{2}{*}{Efficiency}  \\ 
	verb & object     & & & &  & \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
	adposition    &    NP       & 	\emph{to}            & \emph{a friend} \\

copula    &    NP        &     \\
	\emph{	is}        & \emph{a friend} \\
auxiliary    &    VP       &   \\
	has          & written \\
noun    &    genitive      &  \\
friend &  of John \\
noun    &    relative clause       &  \\
	books & that you read \\
complementizer    &    S        & \\
	that & Mary \\
verb    &    PP     &    \\
	went & to school \\
want    &    VP      &    \\
	wants   & to to leave \\
verb    &    subject    &  \\
	(there) entered & a tall man \\
verb    &    manner adverb     & \\
	ran & quickly\\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
    \multicolumn{6}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}
\end{tabular}
	\end{center}
	\caption{Greenbergian Correlations. Following \cite{dryer1992greenbergian}, each correlation is stated in terms of a pair of a `verb patterner' and an `object patterner', whose relative order correlate with that of verbs and objects.
For each correlation, we provide an example.
Given the statistical nature of the correlations, not every language satisfies every one of them.
Not all correlations are satisfied by every natu
}\label{table:corr-dryer}
\end{table*}

\section{Formalizing Efficiency in Word Order}

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/communication.png}
	\caption{Our model of communication: A speaker (left) expresses a dependency structure into a string of words forming a sentence. A listener probabilistically recovers a dependency structure. In this example, the grammar of English allows the listener to recover the correct dependency structure with very high confidence.}
	\label{fig:comm}
\end{figure*}

\begin{figure}
    \centering
    \includegraphics[width=0.45\textwidth]{figures/grammars.png}
	\caption{Given a set of dependency structures, grammars define a set of sentences. For simplicity, we use English words.  }
	\label{fig:pareto-plane}
\end{figure}

We model the process where a speaker communicates an utterance to a listener (Figure~\ref{fig:comm}).

We model the meaning that the speaker communicates via the predicate-argument structure.

We represent these in the format of Dependency Grammar:
This linguistic formalism draws directed arcs between syntactically related words, annotated with syntactic relations.

By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined; the dependency tree (including the labels on the arcs) specifies how the meanings of words are combined.

We can view this as a formalization of the Language of Thought (CITE) TODO explain why

When uttering, the speaker needs to choose an ordering in which to order the words in this tree to generate a string of words.

We are interested in the following hypothesis:
If speakers choose to order their sentences in such a way that efficiency is optimized, will this lead to the observed language universals?

Do languages optimize their word order for efficiency?
Does this optimization explain the observed language universals?

All natural languages have some degree of word order regularities that are specified in their grammar.
It is thus not realistic to assume that speakers optimize each utterance individually.
Instead, we hypothesize that the word order rules of languages as a whole are optimized for efficiency, and that this optimization produces the observed universals.

We formalize a language as a distribution over dependency structures $T$ together with an ordering grammar $\theta$ that specifies how dependency structures are ordered into strings of words.
Applying the ordering grammar to the dependency structures, a language defines a probability distribution over sentences $L$.
Efficiency is defined as a tradeoff between

The \key{predictability} of the language is the negative entropy of the sentences:
\begin{equation}
	R_{Pred} := H[L] = \sum_{l} p(l) \log p(l)
\end{equation}
where the sum runs over all possible sentences $l$ that belong to the language.
This quantity describes how homogeneous the language is, i.e., it is larger if the distribution over sentences is concentrated on a smaller number of frequent sentences. %how many distinct sentences there are.
%It is larger

The opposing pressure is \key{parsability}, which is the mutual information beteen sentences and dependency structures:
\begin{equation}
	R_{Pars} := I[L,T] = \sum_{t,l} p(t,l) \log \frac{p(t|l)}{p(t)}
\end{equation}
where the sum runs over all possible pairs of sentences $l$ and dependency structures $t$ in the language.
This quantity describes the degree to which dependency structures can be unambiguously recovered from sentences.
It is large when they can mostly be recovered, and small if sentences have high amounts of ambiguity about the dependency structure.

The \key{efficiency} of a language is a weighted combination
\begin{equation}
	R_{Eff} := R_{Pars} + \lambda R_{Pred}
\end{equation}
with an interpolation weight $\lambda \in [0,1)$. In all experiments in this paper we use $\lambda=.9$ (see SI appendix section 5 for mathematical justification).

%This combined equation represents efficiency: bits of information transferred about dependency trees minus the cost of the transmission. In expectation over utterances (substituting Eq.~\ref{eq:combined-score} into Eq.~\ref{eq:exp-over-utts}), the overall objective function to be maximized by a word order grammar can be written in information-theoretic terms as:

%The two terms of Eq.~\ref{eq:efficiency} implement two opposing pressures. Parseability (the mutual information term) pushes to make strings more heterogeneous, so that they can indicate different tree structures unambiguously. Predictability (the entropy term) pushes to make strings more homogeneous, to reduce the complexity of using the language.
%The idea that natural language arises from a tension between these two forces goes back over a century \cite{gabelentz1901sprachwissenschaft}.
Eq.~\ref{eq:efficiency} is a special case of the objective function proposed in \cite{ferrericancho2002zipf,ferrericancho2007global,futrell2017memory} as a general objective for communicative systems, taking unordered dependency trees $T$ as the underlying meanings to be conveyed. Eq.~\ref{eq:efficiency} can also be seen as a simplified form of the Information Bottleneck \cite{tishby1999information}, a general objective function for lossy compression which has recently been applied to explain linguistic phenomena such as color naming systems \cite{zaslavsky2018efficient} (see SI section 4 for the precise relationship).

We create

Fortunately, we have exacly the kind of data we need

For the dependency structures, we take real tree structures from corpora of 51 languages.
These corpora represent a typologically and genetically diverse group of languages.

For each of these corpora, we construct grammars that optimize efficiency of the linearized sentences.
(The original word orders of the languages do not enter.)

See SI for our method for creating optimized grammars.

%\subsection{Dependency Length Minimization}

\section{Results}

%These results are shown in more detail for two typologically distinct languages (English and Japanese) in Figure~\ref{fig:predict-pars}. Here we include also the predictability and parseability of the original language data and of the word order grammars optimized for dependency length. We see that real languages are even more predictable than any optimized or baseline language, while optimization for either parseability or dependency length suffices to raise parseability to the level of the real languages. This result is in line with the finding that languages with longer dependencies are harder to parse in general \cite{gulordava2016multilingual}.

%\begin{figure*}
%\includegraphics[width=0.98\textwidth]{../results/plot-by-lang/grid-optimized.pdf} % need MLE langs and Efficiency langs on here
%	\caption{Predictability (top) and parseability (bottom) of real, optimized, and random languages, for English (left) and Japanese (right). We provide the entropy $H[L_\theta(T)]$ for predictability (lower is better) and the conditional entropy $H[T|L_\theta(T)]$ for parseability (lower is better). Both measures are normalized by sentence length.}\label{fig:predict-pars}
%\end{figure*}
%

\subsection{Efficiency Explains DLM}
Finally we demonstrate the relationship between dependency length minimization and the maximization of efficiency. Figure~\ref{fig:deplength} shows average dependency length per sentence length for four typologically distinct languages, showing real languages, random baselines, and languages optimized for dependency length, parseability, predictability, and efficiency. We see that optimizing for efficiency lowers dependency length relative to random baselines, in keeping with the suggestion that dependency length minimization is a by-product of efficiency maximization \cite{futrell2017memory}. In 80\% of the languages, optimizing explicitly for dependency length produces dependencies that overshoot the depenndency length of the real language; in 3/4 of the lanuages shown, the real language is best matched by efficiency optimization.

\section{DLM}

We compare Efficiency to the predictions of Dependency Length Minimization.
This theory states that natural languages order information in such a way that the distances between syntactically linked words are minimized.
For instance, in Figure~\ref{fig:arabic-japanese-simple}, there would be syntactic links between the adposition (to) and the noun (friend), and between the verb and both the object (letter) and the adposition (to).
There is strong evidence that natural language minimizes the length of these dependencies~\cite{ferrericancho2004euclidean,liu2008dependency,gildea2010grammars,futrell2015largescale}.
Theoretical work in the functional linguistics literature has proposed that this principle explains the correlations.
However, the principle itself is stipulative and not a first-principles explanation.
\cite{hawkins1994performance} theoetically argues that it increases  parsability.

However, these arguments have been made on a theoretical basis.
We will show that (1) Dependency Length Minimization indeeds predicts most of the correlations, and (2) Dependency Length Minimization is explained by efficiency optimization.

These computational results confirm theoretical ideas  that have been stated informally by authors in the functional linguistics literature at least since the 1980s.

%the meaning

\section{Minimizing Dependency Length Predicts Universals}

..

\begin{figure}
\centering
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
%	   I \& wrote \& the \& letter \& in \& Japanese \& . \\
	   katabt \& risāla \& li \& sadīq  \\
	   \textsc{verb} \& \textsc{noun} \& \textsc{adp} \& \textsc{noun} \\
	   wrote \& letter   \& to \& friend  \\
   \end{deptext}
	%   \deproot{3}{ROOT}
   \depedge{1}{2}{obj}
	%   \depedge[edge start x offset=-6pt]{2}{5}{ATT}
   \depedge{1}{3}{}
   \depedge{3}{4}{}
   %\depedge[arc angle=50]{7}{6}{ATT}
\end{dependency}
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
      tomodachi \& ni \& tegami-o \& kaita \\
 \textsc{noun} \& \textsc{adp} \& \textsc{noun} \& \textsc{verb}\\
	   friend \& to \& letter  \& wrote \\
   \end{deptext}
	%   \deproot{3}{ROOT}
   \depedge{4}{2}{}
	%   \depedge[edge start x offset=-6pt]{2}{5}{ATT}
   \depedge{2}{1}{}
   \depedge{4}{3}{obj}
   %\depedge[arc angle=50]{7}{6}{ATT}
\end{dependency}
\end{figure}

Dependency Length:
